---
title: "Statistical model, calibration, and numerical guarantees"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical model, calibration, and numerical guarantees}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(heavilyright)
```

This guide specifies the implemented conventions. It is not a proof that every
user-supplied study model satisfies them. A successful numerical calculation
does not validate independence, normality, or a chosen covariance estimator.

## Study model and covariance of the mean

For study $j$, let $y_j$ have dimension $q_j$, let $\theta$ have dimension $d$,
and let $A_j$ map the common parameter to the study estimand. Supply
`study(estimate = y, vcov = V, projection = A, df = nu)` with

\[
  E(y_j) = A_j\theta, \qquad
  Q_j(\theta) = (A_j\theta-y_j)^T V_j^{-1}(A_j\theta-y_j).
\]

`V` is the covariance of the **estimate**. For $n$ independent observations,
the implemented sample-mean covariance is

\[
  \widehat V = \frac{W}{n(n-1)}, \qquad
  W=\sum_{i=1}^n(x_i-\bar x)(x_i-\bar x)^T.
\]

This equals the Bessel-corrected sample covariance divided by $n$, not
$W/n^2$. Its diagonal agrees with the squared componentwise standard errors.

```{r covariance-contract}
x <- matrix(c(1, 2, 4, 7, 2, 5, 8, 11), ncol = 2)
V <- covariance_of_mean(x)
stopifnot(isTRUE(all.equal(V, cov(x) / nrow(x))))
stopifnot(isTRUE(all.equal(diag(V), standard_error_of_mean(x)^2)))
V
```

With known covariance, `df = NULL` uses
$p_j(\theta)=P(\chi^2_{q_j}\ge Q_j(\theta))$. For a sample mean with
multivariate normal observations, use the sample-mean covariance above and
$\nu_j=n_j-1$. Hotelling calibration uses

\[
  p_j(\theta) = P\left(F_{q_j,\nu_j+1-q_j} \ge
    \frac{\nu_j+1-q_j}{q_j\nu_j}Q_j(\theta)\right).
\]

Supplying `df` does not make an arbitrary estimated covariance a Hotelling
covariance. In particular, unequal arm sample sizes in a treatment trial do
not automatically determine a valid common Hotelling degree of freedom.
For $q=1$, this formula gives the usual two-sided Student-t p-value.

```{r scalar-hotelling}
radius <- 1.2
nu <- 19
stopifnot(isTRUE(all.equal(
  pf(radius^2, 1, nu, lower.tail = FALSE),
  2 * pt(radius, nu, lower.tail = FALSE)
)))
```

## Combination and confidence-region inversion

For $m>1$, positive fixed weights $w_j$ sum to one. The two convex-region
methods use

\[
  G(\theta)=\sum_{j=1}^m w_j\phi(p_j(\theta)), \qquad
  \phi_{\mathrm{HC}}(p)=\cot(\pi p/2), \quad
  \phi_{\mathrm{EH}}(p)=1/p.
\]

Small p-values yield large scores. Under independent uniform null p-values
$U_j$, `calibration = "finite"` numerically evaluates the law of
$\sum_j w_j\phi(U_j)$. It uses its upper-tail probability for the combined
p-value and its $(1-\alpha)$ quantile $c_\alpha$ for rejection. The closed
confidence region is

\[
  C_{1-\alpha}=\{\theta:G(\theta)\le c_\alpha\}.
\]

The boundary convention has no probability effect under a continuous null.
Finite-number calibration describes the finite **number of combined tests**;
it is distinct from the finite-sample covariance and Hotelling calibration.
Independence is required for the stated finite-number law. Dependent p-values
need a separate validity argument; neither convexity nor a solver certificate
provides one. Exact nominal coverage further requires correct study p-values.

```{r finite-law}
rule <- new_combination("hcauchy", weights = c(0.2, 0.3, 0.5),
                        calibration = "finite")
rule$cutoff
phcmean(rule$cutoff, rule$weights, lower.tail = FALSE, diagnostics = TRUE)
```

For HCauchy/EHMP, `"auto"` resolves to finite calibration through 1,000 studies
and to Landau above that count. HMP uses the reciprocal score with Landau
calibration even if `"finite"` was requested. Inspect the returned
`specification$calibration`; for an analysis protocol, specify the choice
explicitly. No automatic branch establishes the accuracy of an asymptotic
approximation for a particular weight vector.

For entropy $H=-\sum_j w_j\log w_j$ and Euler's constant $\gamma$, the Landau
approximation uses location $(2/\pi)(H+1-\gamma)$ and scale $1$ for HCauchy,
and location $H+1-\gamma$ and scale $\pi/2$ for EHMP/HMP. The exported Landau
functions transform their argument to the base approximation as

\[
  z=(x-\text{location})\frac{\pi}{2\,\text{scale}}
    +\log\left(\frac{\pi}{2\,\text{scale}}\right).
\]

Do not assume another package's Landau location/scale convention is identical.
Quantiles invert this package's CDF rather than interpolating a lookup table.

The p-value API also implements Cauchy, Levy, Fisher, Stouffer, Bonferroni,
and Simes rules. Those choices do not inherit the HCauchy/EHMP multivariate
convexity guarantee. Fisher uses equal contributions. Bonferroni and Simes do
not use the supplied weight magnitudes. With one study, all rules use the
internal score `-p` and cutoff `-alpha`, preserving the original study test.

## Convex inference and support intervals

All covariance matrices must be positive definite and the stacked projection
matrices must identify all $d$ parameters. The implemented multi-study convex
regimes are:

| Study calibration | HCauchy | EHMP |
|:--|:--|:--|
| Known covariance | All supported dimensions | All supported dimensions |
| Hotelling, $q=1$ | $\nu\ge1$ | $\nu\ge1$ |
| Hotelling, $q=2$ | $\nu\ge2.5$ | $\nu\ge2$ |
| Hotelling, $q\ge3$ | $\nu\ge q+1$ | $\nu\ge q$ |

For a nonzero direction $a$, `support_interval(fit, a)` returns

\[
  L_a=\inf_{\theta\in C}a^T\theta, \qquad
  U_a=\sup_{\theta\in C}a^T\theta.
\]

These are projections of the entire joint region, not line slices with other
coordinates fixed. On the event that the true parameter lies in $C$, every
such projected interval contains its true contrast. Thus simultaneous coverage
comes from the joint region and its assumptions, not from additional pairwise
adjustment. `coef(fit)` returns the minimum-score point, not necessarily the
midpoint of each projected interval or the GLS estimate.

For one study, let $H=A^TV^{-1}A$, let
$\theta_0=H^{-1}A^TV^{-1}y$, and let $Q_{\min}=Q(\theta_0)$. The ellipsoid
uses the appropriate chi-square or Hotelling cutoff $r^2$ and gives

\[
  a^T\theta_0\ \mathbin{\pm}
  \sqrt{(r^2-Q_{\min})\,a^TH^{-1}a}.
\]

```{r analytic-support}
y <- c(0.2, -0.1)
V <- matrix(c(0.04, 0.01, 0.01, 0.09), 2)
a <- c(1, -1)
fit <- fit_meta_md(list(study(y, V)))
interval <- support_interval(fit, a)
radius <- sqrt(qchisq(0.95, 2) * drop(crossprod(a, V %*% a)))
expected <- sum(a * y) + c(-radius, radius)
stopifnot(interval$success,
          isTRUE(all.equal(c(interval$lower, interval$upper), expected)))
interval
```

Multiple-study calculations whiten residuals by covariance and use translated,
scaled parameter coordinates. Scalar residuals have genuine cusps; auxiliary
absolute-residual variables and bounded subgradients preserve them. SLSQP
provides candidates, followed by Newton refinement and, when needed, COBYLA
and bounded active-manifold attempts. Disconnected parameter components are
normalized and minimized separately.

For disconnected components, writing $G=\sum_j a_jG_j$ with normalized local
scores means each local minimum-gap bound contributes $a_j\Delta_j$ to the
parent bound. Passing each local tolerance does not imply that their sum
passes the parent tolerance. The implementation retains natural-score scaling
and, when necessary, assigns tighter absolute caps in parent score units.
Only loose components are retried, using their remaining solver budgets.
Reconstruction/roundoff allowances and the unchanged aggregate tolerance are
checked after refinement. A failed refinement cannot establish emptiness.

The minimum is checked using a convex tangent lower bound over an enclosing
sublevel domain and a minimum-value gap. At a smooth upper support point the
boundary condition holds and $\nabla G=\eta a$ for $\eta>0$; cusp points use
the corresponding subgradient condition. Endpoint success requires a boundary
residual at most `2e-8` of the remaining feasible score budget and a normalized
KKT residual at most `2e-6`, followed by reconstruction checks in original
units. The one-study analytic route has its own quadratic and score checks.

These are floating-point residual certificates, **not** interval-arithmetic
enclosures or a universal convergence guarantee. Optimizer termination alone
is insufficient. Check `success` and retain both endpoint diagnostics.
Dense Newton refinement is capped at 250 active coordinates; the package does
not currently provide a sparse large-dimensional solver.

## NMA and DAC interpretation

NMA contrast estimates mean `treat1 - treat2`. The global reference fixes one
treatment effect to zero; projection rows encode the remaining differences.
For independent arm means sharing a local baseline with variance $v_0$,
contrast covariance is
$\operatorname{diag}(v_1,\ldots,v_q)+v_0\mathbf{1}\mathbf{1}^T$.
This is constructed by `network_arm_studies()`. Explicit covariance inputs
must use a nonredundant contrast basis. Trials, not correlated rows within a
trial, are the natural independent blocks. No random-effects heterogeneity or
treatment-ranking model is fitted.

DAC means **divide-and-combine over parameter coordinates**. All rows remain
in every selected coordinate block. Blocks partition the columns without
padding an uneven remainder; each contributes its own mean, covariance of the
mean, and `df = n - 1`. Partitioning correlated coordinates does not make their
p-values independent. Use a justified independent-block design or assess the
consequences of dependence separately.

## Empty sets and numerical limits

A minimum above the cutoff is not alone enough to establish emptiness: a
validated convex lower bound must also exceed the cutoff. Study removal is
off by default. An explicit `empty_set_policy()` ranks individual p-values at
the validated minimum, breaks ties by original input index, and removes the
first eligible whole block. Multivariate deletion must preserve projection
rank. Retained weights are renormalized and the cutoff is recomputed using the
resolved calibration after each removal. Adaptive removal is sensitivity
analysis; the original coverage statement does not automatically survive it.

Finite-number laws retain characteristic-function inversion for upper-tail
calibration and use a separate lower-tail representation. Requested quadrature
tolerances are `rel.tol = 2e-11` and `abs.tol = 2e-13`, with at most 600
subdivisions per integral. Integration status, estimated error, and probability
range are checked. Error estimates stay on the unlogged scale, even when a log
probability is requested; they are not rigorous total-error bounds.

### Lower tails without subtractive cancellation

For a reciprocal uniform score $Y=1/U$, shift to $Z=Y-1$ before inversion.
Its Laplace transform for $\Re(s)>0$ is

\[
  L_{\mathrm{EH}}(s)=1-s e^s E_1(s)
    =\int_0^\infty\frac{t e^{-t}}{s+t}\,dt.
\]

For a standard half-Cauchy score the transform is

\[
  L_{\mathrm{HC}}(s)=
  \frac{e^{-is}E_1(-is)-e^{is}E_1(is)}{i\pi}.
\]

The native R complex E1 routine uses the principal-branch power series and a
continued fraction, not a Python runtime or a numerical derivative.
[DLMF power series](https://dlmf.nist.gov/6.6.E2),
[DLMF continued fraction](https://dlmf.nist.gov/6.9.E1).
For moderate reciprocal-transform arguments, the equivalent expression
$L_{\mathrm{EH}}(s)=e^s E_2(s)$ is evaluated directly with the generalized
exponential-integral fraction, avoiding cancellation even before the
large-argument switch.
[DLMF generalized exponential integral](https://dlmf.nist.gov/8.19.E1),
[DLMF incomplete-gamma fraction](https://dlmf.nist.gov/8.9.E2).
Large arguments use the transforms' inverse-power expansions directly,
avoiding subtraction of nearly equal terms. The reciprocal expansion has 32
terms with absolute truncation remainder bounded by $33!/|s|^{33}$ from
$|s|=50$. The half-Cauchy expansion has 26 terms, with remainder bounded by
$(2/\pi)52!/(|s|/\sqrt{2})^{53}$ from $|s|=60$. These bounds concern truncation,
not accumulated floating arithmetic.

For positive distance $z$ from support, select a positive saddle $a$ satisfying
$\sum_j w_j[-L'(aw_j)/L(aw_j)]=z$. The R root search works in log coordinates.
Writing $P(s)=\prod_j L(w_js)$, the density inversion is

\[
  f(z)=\frac{e^{az}P(a)}{\pi}\frac{a}{\sqrt m}
  \int_0^\infty \Re\left\{
    e^{i a z u/\sqrt m}\frac{P(a+i a u/\sqrt m)}{P(a)}
  \right\}\,du.
\]

For the CDF, multiply the normalized integrand by
$a/(a+i a u/\sqrt m)$ and divide the prefactor by $a$. External factors are
combined in logarithms, and small CDFs are evaluated directly. The saddle
changes the numerical contour, not the target distribution.

Two-coordinate lower tails instead use a finite positive convolution. Larger
systems with at least eight comparable weights first use phase-combined
improper quadrature. Other systems, or rejected direct quadratures, use checked
finite half-period integrals and convergence acceleration, with a 100-cycle
limit. The estimated error includes accumulated quadrature error and stability
of successive extrapolants. This R-specific implementation uses Wynn's epsilon
recurrence; it is not identical to Python's Fourier quadrature backend.
[DLMF convergence acceleration](https://dlmf.nist.gov/3.9).
Unconverged cycles raise an error, rather than supplying an unchecked zero.

Very close to support, positive-convolution simplex expansions have analytic
remainder bounds. Elsewhere, simplex and tilted-density/Chernoff upper bounds
can establish underflow. Such zero shortcuts retain at least the smallest
positive floating-point number as an absolute error allowance. The old
half-Cauchy `location - 2.5` zero shortcut is no longer used.

Reciprocal count inputs denote exact support one. Explicit float vectors are
not silently renormalized; their support is the exact real sum of the supplied
binary entries. A compensated sum computes the support distance without first
rounding that sum. This distinction also survives quantile inversion. In an
extreme boundary calculation, three supplied binary approximations of `1/3`
therefore need not have the same answer at one as count `3`.

```{r repaired-lower-tails}
phcmean(0.01, 10, diagnostics = TRUE)
pehmean(2, 100, diagnostics = TRUE)
stopifnot(phcmean(0.01, 10) > 0, dehmean(1e160, 1) > 0)
```

Numerical errors combine quadrature/extrapolation estimates and floating
allowances; they do not promise arbitrary-tail uniform relative accuracy.
Analytic support and infinite endpoints are handled separately. Very extreme
weight scales, unrepresentable saddles, or slow oscillatory convergence can
still produce an explicit error.

The study-p-value floor is `1e-150`. Clipped-tail minima and convex domains
that can reach this floor cannot authorize empty-set deletion. Near-empty,
ill-conditioned, or unrepresentable inputs can fail explicitly. Refer to
`vignette("troubleshooting", package = "heavilyright")` for error classes,
trace retention, and a minimal reproducible report.
