---
title: "Troubleshooting and reproducible diagnostic reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting and reproducible diagnostic reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(heavilyright)
```

## Start with the input contract

| Symptom | Check before changing the analysis |
|:--|:--|
| Intervals unexpectedly narrow | Supply covariance of estimates, not observations; use `covariance_of_mean(x) = cov(x)/n`. Arm inputs require SEs of means, not SDs. |
| `vcov must be positive definite` | Check units, duplicated contrasts, too few observations, and constant or collinear coordinates. Do not silently add a ridge and claim the same statistical model. |
| Projections do not span the parameter space | Remove unidentified parameters from the scientific model or supply identifying studies. NMA requires a connected treatment network. |
| Repeated trial IDs require covariance | Use `network_arm_studies()` or supply joint covariance blocks with a nonredundant contrast basis. The explicit independent-row option does not repair dependence. |
| Hotelling df rejected | Check study dimension and the convexity restrictions in `?fit_meta_md`. Do not change df merely to pass validation. |
| Weights rejected | Supply one strictly positive weight per study block and normalize to sum one. Multi-arm rows in one joint trial receive one block weight. |
| Wrong sign for an NMA contrast | Estimates and output bounds mean first treatment minus second. Verify labels and the reference. |
| DAC partition rejected | Blocks must cover every coordinate exactly once. They partition columns, not observations. |

For low-level null-law functions, `weights = 3` means three equal weights;
`weights = c(0.2, 0.3, 0.5)` supplies a normalized vector. Those functions allow
zero weights, but combination specifications require strictly positive weights.
Finite density, probability, and quantile calls are scalar; use `vapply()` for
several evaluation points. Batched `combine_p()` input instead puts studies in
the final array dimension.

```{r shapes}
rule <- new_combination("hcauchy", m = 3)
p <- rbind(c(0.01, 0.2, 0.4), c(0.2, 0.5, 0.8))
combine_p(p, rule)
vapply(c(5, 10), phcmean, numeric(1), weights = 3, lower.tail = FALSE)
```

## An empty region is not an optimizer failure

| Result or condition | Meaning | Safe response |
|:--|:--|:--|
| Scalar fit with `empty = TRUE` | The returned confidence set is empty. Equal stored endpoints are a sentinel, not a singleton interval. | Report emptiness and retain the trace before any sensitivity analysis. |
| `heavilyright_empty_region_error` | A multivariate region is certified empty and removal is off. | Inspect study compatibility; an opt-in removal analysis changes the analysis. |
| `heavilyright_empty_resolution_error` | A removal limit or rank constraint prevents resolution. | Report the unresolved empty set and retained study indices. Do not quietly enlarge the policy. |
| `heavilyright_empty_numerical_error` | Multivariate minimum, tail, or recalibration checks leave the decision unknown. | Keep the error and history; unknown does not authorize deletion. |
| `support_interval(...)$success = FALSE` | The proposed bounds failed at least one numerical certificate. | Do not report the bounds as certified. Inspect both endpoint diagnostics. |
| `heavilyright_numerical_error` | A numerical calculation could not be validated. | Inspect the message, cause, and any attached diagnostics. It is not evidence of emptiness. |
| `heavilyright_integration_error` / `heavilyright_root_error` | A finite-law integration or quantile inversion failed. | Preserve method, weights, significance level, and error. Do not silently switch calibration. |

Not every validation error carries diagnostics. Multivariate empty-set errors
carry a record in `error$diagnostics`, including earlier removals if a later
step fails. Other errors may contain a nested `cause`; retain the original
condition object when investigating.

```{r scalar-empty}
scalar <- fit_meta_1d(c(-3, 3), se = 0.5)
scalar$empty
scalar$empty_set_diagnostics$final_state
```

The following captures a known multivariate empty region without converting an
unexpected failure into a successful fit:

```{r capture-empty}
studies <- list(
  study(c(-4, 0), diag(2), label = "left"),
  study(c(4, 0), diag(2), label = "right")
)
result <- tryCatch(
  fit_meta_md(studies),
  heavilyright_empty_region_error = identity
)
stopifnot(inherits(result, "heavilyright_empty_region_error"))
conditionMessage(result)
record <- result$diagnostics
record$final_state
record$trace[[1]]$feasible_set_score_lower_bound
record$trace[[1]]$cutoff
```

If scientifically appropriate, run a **separate, labelled sensitivity analysis**:

```{r sensitivity}
sensitivity <- fit_meta_md(studies,
  empty_policy = empty_set_policy(max_removals = 1, warn = FALSE))
sensitivity$empty_set_diagnostics$final_original_indices
```

Turning warnings off does not turn removal off. Omit `empty_policy` to disable
removal. Removal acts on an entire supplied study block. In a coordinate-only
DAC partition, deleting a block usually removes the only information about
those coordinates; rank preservation will therefore prevent that deletion.
Do not interpret the sensitivity interval as retaining proven post-selection
coverage.

## Save a trace without overwriting evidence

`write_empty_trace()` needs the optional `jsonlite` package and refuses to
overwrite an existing file. Use a new destination for every analysis. This
example uses a temporary file and deletes only that file afterward.

```{r save-trace}
if (requireNamespace("jsonlite", quietly = TRUE)) {
  path <- tempfile(fileext = ".json")
  write_empty_trace(record, path)
  restored <- jsonlite::read_json(path, simplifyVector = TRUE)
  stopifnot(identical(restored$final_state, record$final_state))
  unlink(path)
}
```

## Failed endpoint checks

For a returned interval, examine `lower_diagnostics` and `upper_diagnostics`,
including `message`, `scaled_boundary_error`, `kkt_residual`, and the recorded
optimizer attempts where present. The analytic one-study path uses different
diagnostic fields from the numerical multi-study path.

```{r inspect-endpoint}
fit <- fit_meta_md(list(study(c(0.2, -0.1), diag(c(0.04, 0.09)))))
interval <- support_interval(fit, c(1, -1))
stopifnot(interval$success)
interval$upper_diagnostics
```

Increasing `maxeval` in `support_interval()` or `nma_contrasts()` can help when
an evaluation budget is exhausted. It cannot establish an unsupported
convexity regime, fix missing parameter rank, or remove probability-floor
limitations. It does not change the earlier minimum search in `fit_meta_md()`.

If all component minima passed but their aggregate gap is still unresolved,
the combined error bounds exceeded the parent allowance. The solver attempts
tighter local certificates without increasing that allowance or replenishing
the original component work budget. Inspect `minimum_diagnostics` on the
underlying numerical condition (`result$cause` for a caught fitting error),
including `initial_optimality_gap_bound`, `component_refinement_attempts`, and
the final gap/tolerance. Exhausted or rejected retries remain numerical
uncertainty and do not authorize study removal.

Rescaling extreme measurement units and expressing a contrast with a smaller
direction magnitude may improve representability, but transform estimates,
covariances, projections, and the final contrast consistently.

`contains()` uses the literal fitted cutoff. A boundary point may be a few
floating-point units outside that cutoff while meeting its numerical boundary
tolerance. Inspect the endpoint's certificate rather than replacing the
returned point or treating membership alone as a solver-quality test.

## Calibration and unexpected differences

Record the resolved calibration, not just the requested `"auto"` setting.
Finite and Landau results need not agree in the last digits or at small study
counts. After removal, weights are renormalized and the cutoff is recalculated
using the resolved calibration. Refit scalar models to change confidence level;
`confint(fit, level = ...)` does not recalibrate an existing fit. Likewise,
`combination_cutoff(alpha, rule)` does not mutate `rule`.

`diagnostics = TRUE` in finite null-law density/probability functions reports
an unlogged numerical error estimate, not a confidence interval and not a
log-scale error. Analytic support boundaries and bounded underflow may return
zero; underflow has a positive error allowance. Tiny lower-tail CDFs are
calculated directly, so use `lower.tail = TRUE` instead of subtracting a
survival probability from one. Numerical failure is never a substitute zero.

An integer reciprocal count has support exactly one. An explicit vector keeps
its binary floating-point values without renormalization, so its exact support
can differ slightly from a rounded sum. This matters extremely close to one;
use the count input when a mathematical equal-weight mean is intended.

For R/Python comparisons, match the covariance divisor, degrees of freedom,
study block grouping, treatment signs, weights, calibration, and empty-set
policy first. Compare endpoint residuals and statistical meaning as well as
numbers; optimizer iteration counts and random draws are not expected to match.

## A useful issue report

Provide a small synthetic example, the complete call, expected and observed
behavior, and the exact condition message. Include `sessionInfo()`, package
versions, resolved calibration, support diagnostics, and an empty-set trace
when available. Do not include identifiable study or participant data.

```{r versions}
packageVersion("heavilyright")
packageVersion("nloptr")
```

For documentation discovery, run `help(package = "heavilyright")` and
`vignette(package = "heavilyright")`. If vignettes are missing after a source
installation, install from a source archive built with vignettes, or rebuild
locally with the suggested `knitr` and `rmarkdown` packages and Pandoc available.
Runtime computation needs neither Python nor the vignette-building tools.
