---
title: "Network meta-analysis, coordinate blocks, and empty regions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network meta-analysis, coordinate blocks, and empty regions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(heavilyright)
```

## Treatment networks

Positive contrasts mean the first treatment minus the second. Treatment names
and the reference are user supplied. A two-treatment network is also supported.

```{r}
network <- network_studies(
  c(0.2, 0.25, 0.1), c(0.2, 0.18, 0.22),
  treat1 = c("B", "C", "C"), treat2 = c("A", "A", "B"), reference = "A"
)
fit <- fit_nma(network)
coef(fit)
contrasts <- nma_contrasts(fit)
contrasts
attr(contrasts, "intervals")[[1]]$upper_diagnostics$kkt_residual
```

For multi-arm trials, independent arm means induce correlated contrasts through
their shared local baseline. Supply standard errors of the arm means.

```{r}
network <- network_arm_studies(
  mean = c(1, 1.2, 0.8, 1, 1.15), se = rep(0.1, 5),
  treatment = c("A", "B", "C", "A", "B"),
  study_id = c("trial1", "trial1", "trial1", "trial2", "trial2"),
  reference = "A"
)
network$studies[[1]]$vcov
nma_contrasts(fit_nma(network))
nma_wls(network)
```

Alternatively pass nonredundant trial contrasts and their joint covariance
matrices to `network_studies()`. Repeated trial identifiers without covariance
raise an error. The explicit `multiarm = "independent"` choice uses a
comparison-level HCCT model; it does not correct within-trial
dependence. These are common-effect regions; no heterogeneity model is fitted.

## Divide-and-combine

DAC divides the parameter coordinates while keeping every observation row.
Unequal or named groups are allowed; the groups must partition all columns.

```{r}
set.seed(21)
x <- matrix(rnorm(120), 30, 4, dimnames = list(NULL, letters[1:4]))
blocks <- list(first = c("a", "c"), second = c("b", "d"))
fit <- fit_dac(x, blocks = blocks)
coef(fit)
contains(fit, colMeans(x))
support_interval(fit, c(1, 0, 0, 0))
```

Every block uses `cov(x)/n` and df `n-1`. These finite-sample calibrations do
not make correlated coordinate blocks independent. The chosen finite-number
combination law assumes independent block p-values; general dependence may
change coverage. `study()` and `fit_meta_md()` provide the lower-level route
for custom estimators, supplied covariance summaries, or overlapping projections.

## Empty-set sensitivity analysis

Adjustment is disabled by default. An opt-in policy removes whole input study
blocks only after a convex lower bound proves emptiness and only if the
remaining projections retain full parameter rank.

```{r}
studies <- list(
  study(c(-4, 0), diag(2), label = "left"),
  study(c(4, 0), diag(2), label = "right")
)
fit <- fit_meta_md(studies, empty_policy = empty_set_policy(warn = FALSE))
fit$empty_set_diagnostics$final_original_indices
fit$empty_set_diagnostics$trace[[1]]$feasible_set_score_lower_bound
fit$empty_set_diagnostics$trace[[1]]$cutoff
support_interval(fit, c(1, 0))
```

The audit retains every ranked candidate, rank check, removal, recalibration,
and numerical failure. Save it with `write_empty_trace(record, new_path)`.
Adaptive removal is a sensitivity procedure, not a post-selection coverage
guarantee. Unrepresentable tails or failed numerical certificates never
authorize deletion. Always check `support_interval(...)$success` before using
the numerical bounds; `nma_contrasts()` raises if a contrast fails its checks.
